# Supplement to Chapter 18

This chapter implements much of the analysis shown in chapter 20 of 
A Modern Introduction to Probability and Statistics. R code is given for the
simple textbook datasets used in the book, and then the concepts are
illustrated on real data.

All datasets from the book can be downloaded here: https://www.tudelft.nl/en/eemcs/the-faculty/departments/applied-mathematics/applied-probability/education/mips/.

This week's assigned exercises are from MIPS, as follows: 18.1 18.2; 18.3, 18.4; 18.6; 18.7; 18.8; 18.9; 18.11. Answers to selected exercises are in the 
back of the book. You should also do all the "quick exercises" within chapter 18 
(solutions are at the end of the chapter). Use `R` as much as possible when answering the
questions.

## The Bootstrap (Chapter 18)

The bootstrap is one of the most foundational computational tools in modern
statistics. It as important on the computation side as basic calculus is on the
math side. Any time you have an estimator and are interested in its repeated
sampling properties (which is much of what statistics *is*), you can simulate
data, compute the estimator, and just *look* at the results to get an idea
of what's going on. This is called the "parametric" bootstrap, and you have
been doing it for the whole course in these supplementary notes.

This chapter introduces the "empirical" bootstrap, which allows you to use this
idea when you don't know the distribution of the data! The idea is to sample
with replacement from the data that you have. A lot of the statistical properties
of estimators are still recoverable under this type of sampling.

It seems like magic. Indeed, the name "bootstrap" comes from the old expression
"lift yourself up by your bootstraps", which roughly means to do something 
seemingly impossible or contradictory in order to better your situation. One
of the best-named concepts in all of stats, in my opinion.

### Empirical bootstrap: Old Faithful data

Let's recreate Figure 18.1:

```{r faithful-1}
library(tidyverse)
# Read in old faithful, copied from chapter 1:
oldfaithful <- readr::read_csv(
  file = "data/MIPSdata/oldfaithful.txt", # Tell it where the file is
  col_names = "time", # Tell it that the first row is NOT column names, and at the same time, tell it what name you want for the column.
  col_types = "n" # Tell it that there is one column, and it is "numeric" (n)
)
# Check what was read in using the dplyr::glimpse() function
dplyr::glimpse(oldfaithful)

# Bootstrap: sample with replacement
set.seed(45356) # So you can reproduce my results
B <- 1000 # Number of bootstrap resamples to take
n <- nrow(oldfaithful) # Sample size
bootstrapmeans <- 1:B %>%
  map(~sample(oldfaithful$time,n,replace = TRUE)) %>%
  map(~(mean(.x) - mean(oldfaithful$time))) %>% # Compute the CENTRED mean
  reduce(c)

# Plot the hisrogram and density estimate together
tibble(x = bootstrapmeans) %>%
  ggplot(aes(x = x)) +
  theme_classic() +
  geom_histogram(aes(y = ..density..),bins = 20,colour = "black",fill = "transparent") +
  geom_density() +
  scale_x_continuous(breaks = seq(-18,18,by=6)) +
  coord_cartesian(xlim = c(-18,18))
```

**Exercise**: compute the book's estimate of $P(|\bar{X}_{n} - \mu|)$. Using the
above `set.seed(45356)` I got the following (slightly different than the book):

```{r faithful-2,echo=FALSE}
mean(abs(bootstrapmeans) > 5)
```

**Exercise**: now, repeat the entire bootstrap procedure $1000$ times and
plot a histogram and kernel density estimate of $P(|\bar{X}_{n} - \mu|)$.
I got the following with `set.seed(8768432)`:

```{r faithful-3,cache = TRUE,echo = FALSE}
set.seed(8768432)
compute_bootstrap_sample <- function(B = 1000) {
  n <- nrow(oldfaithful) # Sample size
  1:B %>%
    map(~sample(oldfaithful$time,n,replace = TRUE)) %>%
    map(~(mean(.x) - mean(oldfaithful$time))) %>% # Compute the CENTRED mean
    reduce(c)
}
compute_bootstrap_prob <- function(samp) {
  mean(abs(samp) > 5)
}

bootstrapprobs <- 1:B %>%
  map(~compute_bootstrap_sample(B = 1000)) %>%
  map(compute_bootstrap_prob) %>%
  reduce(c)

tibble(x = bootstrapprobs) %>%
  ggplot(aes(x = x)) +
  theme_classic() +
  geom_histogram(aes(y = ..density..),bins = 20,colour = "black",fill = "transparent") +
  geom_density() +
  scale_x_continuous(breaks = seq(0,1,by=.01)) +
  coord_cartesian(xlim = c(.18,.28))
```

### Parametric Bootstrap: software data

Read in the software data. Try to do it yourself before looking:

```{bash software-1}
head data/MIPSdata/software.txt
wc -l data/MIPSdata/software.txt
```

```{r software-2}
software <- readr::read_csv(
  file = "data/MIPSdata/software.txt",
  col_names = "time",
  col_types = "n"
)
glimpse(software)
```

Plot the CDF of the estimated distribution against the empirical CDF of the data.
The $t_{ks}$ described in the book is the maximum vertical distance between these
two curves:

```{r software-3}
lambdahat <- 1/mean(software$time)
samppoints <- seq(0,3000,by=1)
expcdf <- pexp(samppoints,rate = lambdahat)
empcdf <- ecdf(software$time)

tibble(x = samppoints,y = expcdf) %>%
  ggplot(aes(x = x,y = y)) +
  theme_classic() + 
  geom_line() +
  stat_function(fun = empcdf,linetype = "dotdash")
```

Compute a bootstrap sample of $T_{ks}$:

```{r software-4,cache = TRUE}
set.seed(97876856)
compute_tks <- function() {
  samppoints <- seq(0,3000,by=1)
  n <- nrow(software)
  samp <- rexp(n,rate = lambdahat)
  ecdfboot <- ecdf(samp)(samppoints)
  expcdf <- pexp(samppoints,rate = lambdahat)
  max(abs(ecdfboot - expcdf))
}
B <- 1000
tksboot <- 1:B %>% map(~compute_tks()) %>% reduce(c)
tibble(x = tksboot) %>%
  ggplot(aes(x = x)) +
  theme_classic() +
  geom_histogram(aes(y = ..density..),bins = 20,colour = "black",fill = "transparent") +
  geom_density() +
  scale_x_continuous(breaks = c(seq(0,.2,by=.05),.176)) +
  coord_cartesian(xlim = c(0,.2))
```

**Exercise**: the above `compute_tks()` is a bit tricky. Write out the exact algorithm
I use to compute $T_{ks}$. Basically I am asking you to expand on step 2. of the
algorithm presented in the book on page 279.

### Extended example: the standard error of a proportion

Dozens of news articles are published every day with claims that "xx% of Canadians
say that yy", where xx is some percentage and yy is some claim that the news
company wants to make. These articles have a significant impact on our public
discourse.

How could it be that every day, there are so many earth-shattering claims? It's
almost like most of them aren't actually true. But then why are they published,
and why do we believe them?

The missing ingredient is *variability*. These news articles most often only
report *point estimates*-- the "xx%". They don't report the *standard error* of
these estimates, which give you a measure of the variability, and hence uncertainty,
in the conclusions that are made based off of them.

We can use the bootstrap to get an idea of the variability in estimates based off
of surveys. Because most news articles don't link to the raw data (another reason
not to believe the stuff that these for-profit companies push), we'll use a general
budget survey from the Government of Canada, with responses from Alberta residents.
You can get the data in excel format [https://open.canada.ca/data/en/dataset/2a5992c4-a050-46b8-b735-641cbc799a8e](here).
You should open the file in excel and save it as a .csv, or if you don't have excel,
I posted it on the course website for you.

Let's read the data into `R`. We know it has a header row (because we opened it
in excel). There is a numeric ID column, two date columns (which we don't need),
and 58 numeric response columns. I removed the first three (blank) and fourth (information)
rows from my
.csv, and you should do the same in excel if you have it.

```{r budget-1,cache = TRUE}
col_types <- c(
  "ncc",
  stringr::str_c(rep("n",58),collapse = "")
) %>%
  stringr::str_c(collapse = "")
col_types # Understand the above code by running it in pieces and looking at the results.

budget <- readr::read_csv(
  file = "data/budget.csv",
  col_names = TRUE,
  col_types = col_types
)
glimpse(budget)

```

Ugly! There are some missing column names that were in the fourth row. If we needed
all these data, we would have to go in excel and manually label the columns correctly,
since the analysts at the government didn't.

Look at the column names and start to think about the kind of news headlines they could
generate. I like the first one: "Q1: To what extent do you think low oil prices impact the Alberta government's ability to budget?" could lead to things like "Albertans have given up hope: more than half 
of Albertans believe their province is unable to budget due to the low price of oil." Hey,
it generates clicks!

Here, let's estimate the proportion of Albertans who answered $1$ ("A Great Deal")
to the above question.

**Exercise**: there are `nrow(budget)` survey responses. The number who respond 1
can be modelled as a Binomial random variable. Derive and calculate the standard
deviation of the sample proportion of respondents who answered 1. I got $0.0428$.
Is this higher or lower than you would expect with this number of respondents?

```{r budget-2}
set.seed(29394032)
origsamp <- as.numeric(budget[["Q1: To what extent do you think low oil prices impact the Alberta government's ability to budget?"]] == 1) 
n <- length(origsamp)
# Convert to 0/1 indicator of response == 1
# Sample proportion:
phat <- mean(origsamp)
phat
# Standard error (you should DERIVE this formula):
sqrt(phat*(1-phat)/n)
# So the mean is .5378 and the standard error is about 0.0428 (theoretically).
# Bootstrap: repeatedly resample and calculate the mean. The estimated standard error
# of the sample mean is then the standard error of these means.
doboot <- function(B) {
  1:B %>%
    map(~sample(origsamp,n,replace = TRUE)) %>%
    map(mean) %>%
    reduce(c)
}
bootmeans <- doboot(1000)
mean(bootmeans) # Pretty close!
sd(bootmeans) # Not bad at all!

tibble(x = bootmeans) %>%
  ggplot(aes(x = x)) +
  theme_classic() +
  geom_histogram(aes(y = ..density..),bins = 30,colour = "black",fill = "transparent") +
  stat_function(fun = rlang::as_function(~n * dbinom(round(n * .x),size = n,prob = phat))) +
  labs(title = "Bootstrap distribution of the sample mean",
       subtitle = "Curve: theoretical distribution")


```

**Exercise**: inside `stat_function(fun = rlang::as_function(~n * dbinom(round(n * .x),size = n,prob = phat)))` I
specify the theoretical density of the sample mean. Derive this formula yourself using the change of variables
formula.

**Exercise**: derive an appropriate *normal approximation to the binomial distribution* for use in this example.
Add it to the above plot using another `stat_function` call, with `colour = "red"` and `linetype = "dotdash"`. I got the following:

```{r budget-3,echo = FALSE}
tibble(x = bootmeans) %>%
  ggplot(aes(x = x)) +
  theme_classic() +
  geom_histogram(aes(y = ..density..),bins = 30,colour = "black",fill = "transparent") +
  stat_function(fun = rlang::as_function(~n * dbinom(round(n * .x),size = n,prob = phat))) +
  stat_function(fun = dnorm,args = list(mean = phat,sd = sqrt(phat*(1-phat)/n)),colour = "red",linetype = "dotdash") +
  labs(title = "Bootstrap distribution of the sample mean",
       subtitle = "Curve: theoretical distribution")
```

The normal approximation looks pretty good!

**Exercise**: suppose that based on this survey, a newspaper claims that "Albertans have given up hope: more than half 
of Albertans believe their province is unable to budget due to the low price of oil.".
Use the bootstrap to calculate the probability that their claim is true, by considering
whether each resampled sample mean is greater than .5. Using `set.seed(5647244)`, and
$B = 1000$ resamples, I got the following:

```{r budget-4,echo = FALSE}
set.seed(5647244)
B <- 1000
1:B %>%
  map(~sample(origsamp,n,replace = TRUE)) %>%
  map(~as.numeric(mean(.x) > .5)) %>%
  reduce(c) %>%
  mean()

```

What do you think about this? Is the newspaper's claim substantiated by the data?

**Exercise**: surveys usually aren't this big. Repeat the above calculation but
on a subset of only $n = 200$ people. What do you get now? Are you going to check
the sample size next time you read a news article that reports survey results?
Use `newsample <- sample(origsamp,200,replace = FALSE)` and then repeate the 
bootstrapping on the `newsample`. Use the same random seed.

```{r budget-5,echo = FALSE}
set.seed(5647244)
newsample <- sample(origsamp,200,replace = FALSE)
n <- length(newsample)
B <- 1000
1:B %>%
  map(~sample(newsample,n,replace = TRUE)) %>%
  map(~as.numeric(mean(.x) > .5)) %>%
  reduce(c) %>%
  mean()
```
