```{r load-tidy-123266341,message=FALSE,warning=FALSE}
library(tidyverse)
```

# Extended Example: World Population Data

In this chapter, we analyze the United Nations' [World Population Prospects](https://population.un.org/wpp/Download/Standard/Population/) data. These
data contain estimated population sizes for each country for single years from
1950--2020, and projections for five year blocks from 2025--2100. 

We will apply methods from each chapter of this course to answer questions
about these data, the underlying true population of the world, and what the 
population will look like in the future. More specifically, we will:

1. Read in and prepare the data for analysis (**Chapter 2**),

1. Investigate the properties of the average population of the worlds' countries
(**Chapter 3**),

1. Build a simple linear regression model for the total world population over time
(**Chapter 4**),

1. Derive two estimators for the total world population and contrast them 
on simulated data (**Chapter 5**),

1. Develop a Bayesian model for total world population (**Chapter 6**),

1. Contrast the two estimators for total world population on the real data,
using bootstrapping (**Chapter 7**),

1. Find and do inference for the maximum likelihood estimator for total
world population (**Chapter 8**),

1. Quantify uncertainty in our estimate for total world population (**Chapter 9**),

1. Assess the goodness of fit of our model for total world population (**Chapter 10**),

1. Develop a Bayesian estimator for total world population (**Chapter 11**),

1. Predict the population to the year 2100 and compare our predictions and
uncertainty quantification with those reported by the UN (**Chapter 12**).

## Read in and prepare the data for analysis

The [World Population Prospects](https://population.un.org/wpp/Download/Standard/Population/) data
contain estimated population sizes for each country for single years from
1950--2020, and projections for five year blocks from 2025--2100. They are available
for free download from that link. They are not in any form suitable for analysis,
however. The data are contained in a spreadsheet with variables in both rows and
columns, and summary statistics mixed in with the raw data.

I have done a bit of manual editing and posted the files to the book data folder
at `data/worldpop/*`. Let's read in the data corresponding to the world population
estimates from 1950--2020. First, look at it, either in excel or on the command line:

```{bash printworldpop-1}
head data/worldpop/worldpop-estimates.csv
```

There is one character column containing the country and then $(2020 - 1950 + 1) = 71$
numeric columns containing population counts.

Let's read it in with those specs:
```{r readworldpop-1}
worldpop <- readr::read_csv(
  file = "data/worldpop/worldpop-estimates.csv",
  col_names = TRUE,
  col_types = stringr::str_c(c("c",rep("n",71)),collapse = "")
)
glimpse(worldpop)
```

Does that look correct to you?

No. Why is the world population only 2 for 1950? If you read the documentation
for the data, you may notice that population is recorded in thousands, but I still
think that there were more than $2,000$ people in the world in 1950. Also, I'm 
pretty sure the single country of Comoros shouldn't have more people than the 
entire world. 

Always look at the data when you read it in.

The problem is debugged by printing the data out on the command line like we did above,
and noticing that in the original file, numbers are stored with spaces in them.
We have to remove these spaces for `R` to read in the data correctly. This kind
of simple but annoying thing happens *all the time* when analyzing data "in the wild".

To remove the spaces (this works for any annoying character like a period, or a 
dollar sign, or whatever), read the data in with all columns as character, process
the data in `R`, and then convert to numeric. Check it out:

```{r readworldpop-11}
remove_space <- function(x) stringr::str_remove_all(x," ")
worldpop <- readr::read_csv(
  file = "data/worldpop/worldpop-estimates.csv",
  col_names = TRUE,
  col_types = stringr::str_c(rep("c",72),collapse = "")
) %>%
  # Remove the space
  mutate_at(vars(`1950`:`2020`),remove_space) %>%
  # Convert to numeric
  mutate_at(vars(`1950`:`2020`),as.numeric)
glimpse(worldpop)
```

Good. Verify that a few values of your choosing match their entries in the original
text data.

The data are in wide format, with the "year" variable contained in the columns. We
want the data in long format for analysis, with two variables, `country` and `year`,
and a variable containing the population count.

We can do that:

```{r readworldpop-2}
worldpop <- worldpop %>%
  pivot_longer(
    `1950`:`2020`,
    names_to = "year",
    values_to = "population"
  )
glimpse(worldpop)




  
```

That looks better! Do the following exercises:

**Exercises**:

1. What is the estimated world population in 2020 (remember, the recorded
values are in **thousands** of people. Answer this question in terms of
**number of people**)?

1. What is the estimated world population in Canada in 1975? Use the `filter` function.

1. The total world population should equal the sum of the population in each country.
Check this. Do the following:

    a. Compute the world population by summing the population of each country.
    Use the `filter` function to remove the `WORLD` row from each year. Then use
    `group_by` and `summarize` to sum `population` over `country`. Save the result
    in a dataframe called `worldpop_summed`.
    
    b. Pull the UN's estimated world population by using the `filter` function to
    *keep* only the `WORLD` row from each year. Save this in a dataframe called
    `worldpop_un`.
    
    c. Join them, and make a plot of the *difference* between the sum of the countries'
    populations and the UN's estimate, for each year. Here's what I got for both the
    data and the plot:
    
```{r worldpopexercise-1,echo=FALSE}
worldpop_summed <- worldpop %>%
  filter(country != "WORLD") %>%
  group_by(year) %>%
  summarize(worldpop = sum(population))

worldpop_un <- worldpop %>% filter(country == "WORLD")

worldpop_joined <- inner_join(worldpop_summed,worldpop_un,by = "year")
glimpse(worldpop_joined)
worldpop_joined %>%
  ggplot(aes(x = year,y = worldpop - population)) +
  theme_classic() + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Difference in population estimates",
       x = "Year",
       y = "Sum of countries - UN estimate") +
  scale_x_discrete(breaks = as.character(seq(1950,2020,by=5)))
```

*Hints*: use `scale_x_discrete(breaks = as.character(seq(1950,2020,by=5)))` to get
the five-year axis, and use `theme(axis.text.x = element_text(angle = 90))` to make the axis text
sideways. Use `geom_bar(stat = "identity")` to get the bar plot. Or, make another
type of plot of your choosing! Don't be afraid to have some fun.

Now, for later in this chapter, we're going to need data on the world population projections
from 2020--2100. The median and "95\% intervals" (you'll learn what this means later)
are all stored in seperate files. I'll read in the "median" one and then you'll read
in the "interval" ones and then join them all together.

The prediction data is mostly in the same format as the population estimates. The
numbers are now stored with quotes and with thousands separated by commas, however.
Like when it was stored with thousands separated by spaces, we have to read it in 
as a character, and then process the data in `R` and convert to numeric.

```{bash printmedian}
head data/worldpop/worldpop-pred-median.csv
```

```{r readmedian-1}
remove_comma <- function(x) stringr::str_remove_all(x,",")
worldpop_pred_median <- readr::read_csv(
  file = "data/worldpop/worldpop-pred-median.csv",
  col_names = TRUE,
  col_types = stringr::str_c(rep("c",18),collapse = "")
) %>%
  # Remove the commas. R already removed the quotes.
  mutate_at(vars(`2020`:`2100`),remove_comma) %>%
  # Convert to numeric
  mutate_at(vars(`2020`:`2100`),as.numeric) %>%
  # Pivot to long format
  pivot_longer(
    `2020`:`2100`,
    names_to = "year",
    values_to = "population"
  )
glimpse(worldpop_pred_median)
```

**Exercise**: read in the `worldpop-pred-lower95.csv` and `worldpop-pred-upper95.csv`
datasets into dataframes called `worldpop_pred_lower95` and `worldpop_pred_upper95`,
with the `population` variable named `population_lower95` and `population_upper95`.
Join the three dataframes into a dataframe `worldpop_pred` which looks like this:

```{r readpoppred-11,echo = FALSE}
worldpop_pred_lower95 <- readr::read_csv(
  file = "data/worldpop/worldpop-pred-lower95.csv",
  col_names = TRUE,
  col_types = stringr::str_c(rep("c",18),collapse = "")
) %>%
  # Remove the commas. R already removed the quotes.
  mutate_at(vars(`2020`:`2100`),remove_comma) %>%
  # Convert to numeric
  mutate_at(vars(`2020`:`2100`),as.numeric) %>%
  # Pivot to long format
  pivot_longer(
    `2020`:`2100`,
    names_to = "year",
    values_to = "population_lower95"
  )
worldpop_pred_upper95 <- readr::read_csv(
  file = "data/worldpop/worldpop-pred-upper95.csv",
  col_names = TRUE,
  col_types = stringr::str_c(rep("c",18),collapse = "")
) %>%
  # Remove the commas. R already removed the quotes.
  mutate_at(vars(`2020`:`2100`),remove_comma) %>%
  # Convert to numeric
  mutate_at(vars(`2020`:`2100`),as.numeric) %>%
  # Pivot to long format
  pivot_longer(
    `2020`:`2100`,
    names_to = "year",
    values_to = "population_upper95"
  )

worldpop_pred <- worldpop_pred_median %>%
  inner_join(worldpop_pred_lower95,by = c("country","year")) %>%
  inner_join(worldpop_pred_upper95,by = c("country","year"))

glimpse(worldpop_pred)
```

We'll use these data later in this chapter.

## Investigate the properties of the average population of the worlds' countries
