# Short tutorial on pulling data for Assignment 1

This "chapter" shows you how to use 
the `opendatatoronto` `R` [package](https://sharlagelfand.github.io/opendatatoronto/)
to pull data from the [Toronto Open Data Portal](https://open.toronto.ca/) 
and the `cesR` `R` [package](https://hodgettsp.github.io/cesR/) to pull data from the [Canadian Election Study](http://www.ces-eec.ca/). This
is to support you doing assignment 1 part 2 for STA238, Winter 2021.


## Toronto Open Data Portal

The [Toronto Open Data Portal](https://open.toronto.ca/) is a website maintained
by the City of Toronto full of publicly accessible datasets pertaining to city operations.
Let's pull a dataset on monthly TTC subway station usage counts.

First install and load the `opendatatoronto` package:

```{r installopendatato1}
# Install the package. You only have to run this line once. Uncomment it 
# first (delete the "#" sign before the "install"):
# install.packages("opendatatoronto")

# Now load the package. You have to do this every time you re-open R:
library(opendatatoronto)
# Oh, and always always load the tidyverse:
library(tidyverse)
```

Check out the open data portal: type `browse_portal()` into your
`R` console and hit `Enter`. It should open up the portal website in a web browser.

I searched for "TTC ridership" and the first thing that popped up was a dataset
of TTC subway station usage counts, among other things. You can do this searching right in `R` using the package:

```{r searchto1}
search_packages("TTC ridership")
```

This is convenient because it gives you identifying information. Let's pull the `id`
of the one we want:

```{r pullid1}
# The search_packages("TTC ridership") command returns a dataframe, so you can 
# index it:
search_packages("TTC ridership")[4, ] # we want this one!
idoftheonewewant <- search_packages("TTC ridership")[4,'id']
# This returns a 1 x 1 dataframe. Use pull() to get it as a single character string:
idoftheonewewant <- pull(idoftheonewewant,id)
idoftheonewewant
```

Each dataset on the portal has potentially several "resources" pertaining to it.
This will include the data itself, but also maybe a readme, some metadata, 
different formats, years, etc. Check what's available for our chosen data using the `list_package_resources()`
function:
```{r list1}
list_package_resources(idoftheonewewant)
```

It appears that there is data for multiple years. Grab the id for 2017:

```{r grabid1}
idofthedata <- pull(
  list_package_resources(idoftheonewewant)[5,"id"],
  id
)
idofthedata
```

Now, to actually get the dataset, you can call `get_resource()`:

```{r getresource1}
thedata <- get_resource(idofthedata)
glimpse(thedata)
```

The data appear a bit messy and will have to be cleaned-- see Chapter 2 for detailed
examples of how to do this.

To summarize, to get a dataset:

1. Search for data on a chosen subject of interest using `search_packages()`,

1. Figure out which one you want and grab its `id`,

1. Check what resources are actually available for those data using `list_package_resources`,
and grab the `id` of the dataset you want to download, and

1. Download the data with `get_resource`.


## Canadian Election Study

The [Canadian Election Study](http://www.ces-eec.ca/) is an annual survey of voting
and other preferences and demographics which are thought to pertain to political
behaviour of Canadian voters. It is publicly available and you can get it using the 
`cesR` package.

If you're reading this tutorial and noticing that this section is shorter than
the Toronto Open data one and you're thinking this is the path of least resistance:
be warned that the CES data are often a LOT messier than some of the datasets you'll
find on the Open Data Portal. So you're just kicking the can down the road in terms
of work :)

Install and load the package like before. It's not on the `CRAN` package repository
and cannot be installed using `install.packages`. The [package website](https://hodgettsp.github.io/cesR/)
tells you how to install it directly from github:

```{r installcesR}
# First, if you don't already have it, install the devtools package:
# install.packages("devtools")

# Now use devtools to install the cesR package directly from Github:
# devtools::install_github("hodgettsp/cesR")

# Load it like any other package:
library(cesR)
```

There are many different CES datasets, and they have unique codes. See them with
the `get_cescodes()` function:

```{r getcescodes1}
get_cescodes()
```

A table of what each of these codes means is available on the website.

Pick one, and get it with the `get_ces()` function:

```{r getces1}
get_ces("ces2015_phone")
```

The function `get_ces` assigns an object `ces2015_phone` in your global environment.
The above code therefore looks slightly different to how you would usually see this step.

So this is all it takes to get the data... but when choosing which data to use you
may want to consider how messy these data are:

```{r cesmessy1}
glimpse(ces2015_phone)
```

