--- 
title: "STA238: Probability, Statistics, and Data Analysis"
author: "Alison Gibbs and Alex Stringer"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
output_dir: "docs"
bibliography: [book.bib,packages.bib]
biblio-style: apalike
link-citations: yes
description: "This book represents part of the course materials for STA238 at the University of Toronto"
---

# Descriptive Statistics: What my data is telling me

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

## Introduction

Data only becomes information after it is analyzed. Analyzing data is extraordinarily difficult,
but is the only way to learn meaningfully about the world.

In this book, you will learn about analyzing data. You will learn how to make principled,
measured statements about the data on hand (*descriptive statistics*); how to 
use the data on hand to make statements about the underlying world, and how to
quantify the uncertainty in your statements (*inferential statistics*), and how
to use the data on hand to make judgements, with uncertainty quantification,
about what data you will see next (*predective statistics*).

This chapter focussed on **descriptive statistics**. This loosely refers to the set
of principles and tools that you use to tell a story using available data. The data
may be collected from a formal scientific experiment, or might be from a survey which
relies on a random sample of a population. It could exist already in a big database
at a bank or insurance company, or it could be scraped from the web. In all
cases, the sitation is: you have data, and you want to use it to *tell a story.*

## Descriptive Statistics

Like with usual storytelling, how you *present* the data makes a big difference.
There is no one correct way to analyze any given dataset. You will make judgements,
and assumptions, and you will have to justify and explain them. These judgements
and assumptions can have a profound impact on the story you tell using your data.

For example, consider a famous dataset containing information on smoking and mortality.
The data is available in the `R` package `faraway`. We may load the package and data
and retrieve information on it as follows:

```{r load-smoke}
# install.packages("faraway") # Run this to install the faraway package
library(faraway) # Attach the faraway package
data("femsmoke") # Load the "femsmoke" data
# ?femsmoke # Run this to open the help page for the dataset.
```

Lines in `R` that begin with a pound sign (or "hashtag" for the younger, more hip reader), $\#$,
are comments and are not run. Remove the $\#$ to run the code.

We see from the help page, and associated reference to the paper in the *American Statistician*,
that the data comes from asking women in Newcastle, England, whether they smoke or not, 
and then following up in 20 years to see if they died. We can use these data to
observe any apparent association between smoking and mortality.

To investigate any such associations, we can simply look at the observed mortality rates
for smokers and non-smokers. This is our first **statistic**. What makes it a "statistic"?
Formally, a **statistic** is any summary---or *function*---of observed data. Here we
are going to report the *mortality rate*---the proportion of people who died---for smokers,
and non-smokers. These are both statistics, and reporting these two statistics side
by side is our first example of an **exploratory data analysis**.

```{r mort-rate-1,warning=FALSE,message=FALSE}
# install.packages("tidyverse")
library(tidyverse) # We use a LOT of functions from this package.

# Compute the mortality rate for smokers and non-smokers.
# To do this, create a dataframe containing the numbers of smokers
# and non-smokers
smoker_numbers <- femsmoke %>% # The %>% operator lets you form sequences of operations
  group_by(smoker) %>% # group_by() makes all the following operations happen within groups
  summarize(num_people = sum(y)) # Count the number of people who are smokers and not smokers 
smoker_numbers

# Now, compute the number of people who died out of the smokers and non-smokers
# This looks the same as above, except we now filter() only the people who died.
smoker_numbers_dead <- femsmoke %>%
  filter(dead == "yes") %>% # Retains rows where dead == "yes" only
  group_by(smoker) %>%
  summarize(num_dead = sum(y))
smoker_numbers_dead

# Now, we join these two tables together and compute the mortality rates by group.
smoker_numbers %>%
  inner_join(smoker_numbers_dead,by = "smoker") %>% # Joins rows with the same value of "smoker"
  mutate(mort_rate = num_dead/num_people) # mutate() creates a new variable, which can be a function of the other variables in the dataframe.
```

See anything interesting?

What went wrong? Why are we concluding that smokers have a *lower* mortality rate than
non-smokers? Two possibilites immediately come to mind:

  1. The opposite is really true, however, we got a very unlikely dataset;
  1. We made a mistake in our analysis.
  
We will address the first notion when we discuss **inferential statistics** later
in the course. The second notion is more immediate. Did we make a mistake?

One thing we definitely did was ignore some present information. Specifically,
we also know how old the women were. How can we include this information in our
descriptive statistics? We can compute mortality rates by age:

```{r mort-rate-2,warning=FALSE,message=FALSE}
smoker_numbers_age <- femsmoke %>%
  group_by(smoker,age) %>%
  summarize(num_people = sum(y))

smoker_numbers_age_dead <- femsmoke %>%
  filter(dead == "yes") %>% # Retains rows where dead == "yes" only
  group_by(smoker,age) %>%
  summarize(num_dead = sum(y))

smoker_numbers_age %>%
  inner_join(smoker_numbers_age_dead,by = c("smoker","age")) %>% 
  mutate(mort_rate = num_dead/num_people)
```

Older people are more likely to die within the 20 year followup period. 
However, examining the raw counts of people in each group, we also see that 
*in these data, older people are less likely to smoke* than younger people. So
*in these data*, less smokers died, because less smokers were old, and more
old people died.

But was our first analysis wrong? No. Our first analysis was fine: we computed
the mortality rate in each group. The problem was in the reporting, or the way
we told the story. We didn't provide enough information when we said "the mortality
rate for smokers was lower than for non-smokers". We should have mentioned that
this is *averaging over all age groups*. Even when we include age in the analysis,
we ought to mention the fact that there are a whole lot of variables we *could*
have measured but didn't, and we are implicitly averaging over these too.
